\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-08-04 analysis/paper/paper.Rmd
fixed plot breaks and litrev. names

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, data visualization]
  plain:     [survival analysis, absolute risk, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation. On the other hand, our package \pkg{casebase} provides penalized estimation in the context of parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.

@@

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of step-wise functions. Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.

@@

\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
\textbf{coxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{Flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
\textbf{Flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{RiskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\
\textbf{Rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\
\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
\textbf{Survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\

@@

```{r supportData, eval = eval_cs3}
data(support)
# Change time to years
support$d.time <- support$d.time/365.25

@@

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_step(aes(y = APS, colour = "APS")) +
  geom_step(aes(y = Full, col = "Full")) +
  geom_step(aes(y = SPS, colour = "SPS")) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme
```

@@

cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
  geom_line() +
  xlab("Follow-up time (Years)") +
  ylab("Brier score") +
  labs(color = "Models") +
  # ggtitle("Brier scores over time") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme

@@

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
  geom_line(aes(y = absRisk, colour = Model)) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
  paper_gg_theme
```

@@

reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
  geom_line() +
  xlab("Follow-up time (Years)") +
  ylab("Brier score") +
  # ggtitle("Brier score over time") +
  labs(color = "Models") +
  paper_gg_theme +  
  scale_color_manual(values = q4)

@@

Future work will look at some of the methodological extensions of case-base sampling. First, to assess the quality of the model fit, we would like to study the properties of the residuals (e.g. Cox-Snell, martingale). More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling. The resulting diagnostic tools could then be integrated in this package. Also, we are interested in extending case-base sampling to account for interval censoring. This type of censoring is very common in longitudinal studies, and many packages (e.g. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it. Again, we hope to include any resulting methodology as part of this package.

@@

In conclusion, we presented the \proglang{R} package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models. We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practioners.

>>>>

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, data visualization]
  plain:     [survival analysis, absolute risk, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}

output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{riskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation. On the other hand, our package \pkg{casebase} provides penalized estimation in the context of parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.

@@

Finally, several packages include functions to estimate the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of step-wise functions. Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.

@@

\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\ \bottomrule
\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\ \bottomrule
\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\ \bottomrule
\textbf{CoxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
\textbf{flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\ \bottomrule
\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{riskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\ \bottomrule
\textbf{rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\ \bottomrule
\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\ \bottomrule
\textbf{survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\ \bottomrule

@@

```{r supportData, eval = eval_cs3}

scaleFUNy <- function(x) sprintf("%.2f", x)
scaleFUNx <- function(x) sprintf("%.0f", x)
data(support)
# Change time to years
support$d.time <- support$d.time/365.25

@@

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_step(aes(y = APS, colour = "APS")) +
  geom_step(aes(y = Full, col = "Full")) +
  geom_step(aes(y = SPS, colour = "SPS")) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)
```

@@

cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
  geom_line() +
  xlab("Follow-up time (Years)") +
  ylab("Brier score") +
  labs(color = "Models") +
  # ggtitle("Brier scores over time") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8)

@@

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
  geom_line(aes(y = absRisk, colour = Model)) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=9)
```

@@

reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
  geom_line() +
  xlab("Follow-up time (years)") +
  ylab("Brier score") +
  # ggtitle("Brier score over time") +
  labs(color = "Models") +
  paper_gg_theme +  
  scale_color_manual(values = q4) + scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)

@@

Future work will look at some of the methodological extensions of case-base sampling. First, to assess the quality of the model fit, we would like to study the properties of the residuals (e.g. Cox-Snell, martingale). More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling. The resulting diagnostic tools could then be integrated in this package. Also, we are interested in extending case-base sampling to account for interval censoring. This type of censoring is very common in longitudinal studies, and many packages (e.g. \pkg{SmoothHazard}, \pkg{survival} and \pkg{rstpm2}) provide functions to account for it. Again, we hope to include any resulting methodology as part of this package.

@@

In conclusion, we presented the \proglang{R} package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models. We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners.
